Contribution Title\thanks{Supported by organization x.}

### 2019-10-25 paper/paper.tex
Chase the whitespace! Don't let your dreams be dreams!

\begin{center}
\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
    scale = 1,transform shape]
  \node[state,initial] (1) {};
  \node[state] (2) [right of=1] {};
  \node[state] (3) [right of=2] {};
  \node[state] (4) [right of=3] {};
  \node[state,accepting] (5) [right of=4] {};

@@

\begin{center}
\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
    scale = 1,transform shape]
  \node[state,initial] (1) {$1$};
  \node[state,accepting] (2) [right of=1] {$2$};
  \node[state,accepting] (3) [right of=2] {$3$};

@@

\begin{center}
\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
    scale = 1,transform shape]
  \node[state] (1) {$1$};
  \node[state] (2) [right of=1] {$2$};
  \node[state] (3) [right of=2] {$3$};

@@

We write our syntax definitions in a DSL building on
\emph{syncons}, introduced by Palmkvist and
Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019},
wherein each production is defined separately from each other (one
\syncon{syncon} each).
%
The tool which performs the parsing is implemented in Haskell and
uses an off-the-shelf
implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
of the Earley parsing algorithm
\cite{earleyEfficientContextfreeParsing1970}.

The main difference between the DSL and the formalism in
Section~\ref{sec:parse-time-disambiguation} is that marks are
introduced as separate \syncon{forbid} declarations rather than
being inlined in productions. This allows supporting high-level
convenience constructs, e.g., specifying precedence between
previously defined operators instead of manually inserting marks,
but is also important for composability as ambiguities can be
statically resolved without changing the original definitions.
Other convenience constructs include defining prefix, infix, and
postfix operators with a given associativity.

The running example used in
Section~\ref{sec:parse-time-disambiguation}
(Fig.~\ref{fig:running-example-definition} on
page~\pageref{fig:running-example-definition}) can be defined as
follows:

@@

\noindent Here, we define a syntax type \syncon{Exp} for
expressions, and declare that parentheses can be used to group
expressions. We define a lexical token for integers, and a
\syncon{syncon} for integer literals using this token. A list is
defined as a sequence of zero or more expressions separated by
semi-colons wrapped in square braces. Finally, addition and
multiplication are defined as infix syncons with the expected
precedence rules. Note that we could just as well have replaced
the precedence list with explicit \syncon{forbid} declarations,
``\syncon{forbid mul.left = add}'' and ``\syncon{forbid mul.right = add}''
(cf. the mark $\{a\}$ in the production of $m$ in
Fig.~\ref{fig:running-example-definition}).

@@

\noindent
Na\"{i}vely composing this definition with another, separately
defined language is likely to introduce ambiguities. For example,
if that language also contains infix operators, the precedence
between these and the Orc combinators will be undefined. With
support for resolvable ambiguity, however, we can allow this
ambiguity and let programmers disambiguate expressions using
parentheses. After composing the above definition with a simple
language supporting addition, variables, and function calls
(language definition omitted for brevity), parsing the expression
%
``\ocaml{1 + 2 >x> f(x)}''
%
results in the following error message from our tool:

@@

\begin{itemize}
\item We could decide to forbid the case where two comparisons are
  right next to each other, e.g., \syncon{forbid gt.left = gt} and
  \syncon{forbid gt.right = gt} (where \syncon{gt} is the syncon
  for the greater-than operator).
\item We could change the definition of the parallel combinators
  and define separate lexical tokens for the combinators, e.g,
  \syncon{token Seq = ">[[:lower:]]}\\\syncon{[[:word:]]*>"}. This
  would add just enough whitespace sensitivity to allow separating
  the different cases.
\item We could change the syntax of the combinators to avoid
  clashes with other operators.
\end{itemize}

\noindent
The change in the first alternative is somewhat ad-hoc, but can be done after composition
and would allow us to reuse both language definitions without
modifications. In the latter two alternatives, we lose reuse of
the definitions of Orc combinators, but place no additional
restrictions on the base language. The preferred resolution
strategy is likely to differ between different compositions and
different languages.
%
For example, another unresolvable ambiguity is going to show up if
the base language uses semi-colons, e.g. for sequencing
expressions (this would clash with the ``otherwise'' combinator).
In this case, there is no reasonable way to disambiguate
``\ocaml{e1; e2}'' without changing the syntax of one of the
operations.

@@

In this section, we investigate how to apply dynamic resolvability
analysis to identify ambiguities in the specification of a
real-world language.
%
Using syncons, we have specified a substantial subset of OCaml's
syntax as described in chapters 7 and 8 (base language and
extensions) of the OCaml reference manual~\cite{leroyOCamlSystemRelease2018}.
The syncon definition is just under 700 lines of code, and is
currently complete enough to parse roughly 75\% (1012 out of 1334
files) of the \verb|.ml| files present in the OCaml compiler
itself (we discuss limitations of our tool in the end of this
section).

@@

Danielsson and Norell~\cite{danielssonParsingMixfixOperators2011} give a method for specifying grammars for expressions containing mixfix operators. They allow non-transitive, non-total precedence, and ``feel that it is overly restrictive to require the grammar to be unambiguous.'' Similar to our approach, they do not reject ambiguous grammars, only ambiguous parses. They also introduce a concept of \emph{precedence correct} expressions; expressions where direct children must be related by precedence (children must have higher precedence than parents). This is more restrictive than our approach, e.g., in a language where \verb|'+'| and \verb|'-'| have no defined relative precedence they reject \verb|'1 + 2 * 3'| as syntactically invalid, while we parse it as an ambiguous expression.

@@

Parsing expression grammars \cite{fordParsingExpressionGrammars2004} sidestep the issue of ambiguity by not introducing it in the first place. However, this also loses the potential gains of leaving certain ambiguities. Additionally, since the ordering of productions matter, composition of languages must be ordered, and the interactions between composed languages becomes non-obvious, e.g., merely adding productions may remove previously recognized words from the language, depending on where they are added.

>>>>

\begin{center}
\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
    scale = .9,transform shape]
  \node[state,initial] (1) {};
  \node[state] (2) [right of=1] {};
  \node[state] (3) [right of=2] {};
  \node[state] (4) [right of=3] {};
  \node[state,accepting] (5) [right of=4] {};

@@

\begin{center}
\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
    scale = .9,transform shape]
  \node[state,initial] (1) {$1$};
  \node[state,accepting] (2) [right of=1] {$2$};
  \node[state,accepting] (3) [right of=2] {$3$};

@@

\begin{center}
\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
    scale = .9,transform shape]
  \node[state] (1) {$1$};
  \node[state] (2) [right of=1] {$2$};
  \node[state] (3) [right of=2] {$3$};

@@

We write our syntax definitions in a DSL building on
\emph{syncons}, introduced by Palmkvist and
Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019},
wherein each production is defined separately from each other (one
\syncon{syncon} each).
%
The tool is implemented in Haskell and uses an off-the-shelf
implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
of the Earley parsing algorithm
\cite{earleyEfficientContextfreeParsing1970}.

The main difference between the DSL and the formalism in
Section~\ref{sec:parse-time-disambiguation} is that marks are
introduced as separate \syncon{forbid} declarations instead of
being inlined in productions. This allows supporting
convenience constructs, e.g., specifying precedence between
previously defined operators instead of manually inserting marks,
but is also important for composability as ambiguities can be
statically resolved without changing the original definitions.
Other convenience constructs include prefix, infix, and
postfix operators with a given associativity.
%
The running example used in
Section~\ref{sec:parse-time-disambiguation}
(Fig.~\ref{fig:running-example-definition} on
page~\pageref{fig:running-example-definition}) can be defined as
follows:

@@

\noindent Here, we define a syntax type \syncon{Exp} for
expressions, and declare that parentheses can be used to group
expressions. We define a lexical token for integers, and a
\syncon{syncon} for integer literals using this token. A list is
defined as a bracketed sequence of zero or more expressions separated by
semi-colons. Finally, addition and
multiplication are defined as infix syncons with the expected
precedence rules. Note that we could also have replaced
the precedence list with explicit \syncon{forbid} declarations,
``\syncon{forbid mul.left = add}'' and ``\syncon{forbid mul.right = add}''
(cf. the mark $\{a\}$ in the production of $m$ in
Fig.~\ref{fig:running-example-definition}).

@@

\noindent
Na\"{i}vely composing this definition with another, separately
defined language is likely to introduce ambiguities. For example,
if that language also contains infix operators, the precedence
between these and the Orc combinators will be undefined. With
support for resolvable ambiguity, however, we can allow this
ambiguity and let programmers use parentheses for disambiguation.
After composing the above definition with a simple language
supporting addition, variables, and function calls (full
definition omitted for brevity), parsing the expression
%
``\ocaml{1 + 2 >x> f(x)}''
%
results in the following error message from our tool:

@@

\begin{itemize}
\item We could decide to forbid the case where two comparisons are
  right next to each other, e.g., \syncon{forbid gt.left = gt} and
  \syncon{forbid gt.right = gt} (where \syncon{gt} is the syncon
  for the greater-than operator).
\item We could change the definition of the parallel combinators
  and define separate lexical tokens for the combinators, e.g,
  \syncon{token Seq = ">[[:lower:]]}\\\syncon{[[:word:]]*>"}. This
  would add just enough whitespace sensitivity to allow separating
  the different cases.
\item We could change the syntax of the combinators to avoid
  clashes.
\end{itemize}

\noindent
The first alternative is somewhat ad-hoc, but can be done after composition
and would allow us to reuse both language definitions without
modifications. In the latter two alternatives, we lose reuse of
the definitions of Orc combinators, but place no additional
restrictions on the base language. The preferred resolution
strategy is likely to differ between different compositions and
different languages.
%
For example, another unresolvable ambiguity is going to show up if
the base language uses semi-colons, e.g. for sequencing
expressions (this would clash with the ``otherwise'' combinator).
In this case, there is no reasonable way to disambiguate
``\ocaml{e1; e2}'' without changing the syntax of one of the
operations.

@@

In this section, we investigate how to apply dynamic resolvability
analysis to find ambiguities in the specification of a
real-world language.
%
Using syncons, we have specified a substantial subset of OCaml's
syntax as described in chapters 7 and 8 (base language and
extensions) of the OCaml reference manual~\cite{leroyOCamlSystemRelease2018}.
The syncon definition is just under 700 lines long, and can
currently parse roughly 75\% (1012 out of 1334
files) of the \verb|.ml| files present in the OCaml compiler
itself (we discuss limitations of our tool in the end of this
section).

@@

Danielsson and Norell~\cite{danielssonParsingMixfixOperators2011} give a method for specifying grammars for expressions containing mixfix operators. They allow non-transitive, non-total precedence, and ``feel that it is overly restrictive to require the grammar to be unambiguous.'' Similar to our approach, they do not reject ambiguous grammars, only ambiguous parses. They also introduce a concept of \emph{precedence correct} expressions; expressions where direct children must have higher precedence than parents. This is more restrictive than our approach, e.g., in a language where \verb|'+'| and \verb|'-'| have no defined relative precedence they reject \verb|'1 + 2 * 3'| as syntactically invalid, while we parse it as an ambiguous expression.

@@

Parsing expression grammars \cite{fordParsingExpressionGrammars2004} sidestep the issue of ambiguity by not introducing it at all. However, this also loses the potential gains of leaving certain ambiguities. Additionally, since the ordering of productions matter, composition of languages must be ordered, and the interactions between composed languages becomes non-obvious, e.g., merely adding productions may remove previously recognized words from the language, depending on where they are added.
